A Simple Minibatch Acceptance Test for MCMC

### 2016-05-19 nips_2016.tex
More revisions to experiment. Revised one figure, fixed some stuff in appendix on proofs, and slight adjustments to labeling figures

We conduct three sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze the efficiency of our sampler on logistic regression. In
Section~\ref{ssec:nets}, we show that our minibatch MH can obtain good MAP estimates for deep
learning.

@@

From the figures above, we find that with the help of our mini-batch MH testing we can obtain the
sampled distribution which are very close to target one. We feature draw our estimation of standard
derivation of $X_{\rm norm}$ over iterations, see Fig.~3e. We find with even constant step size, the
standard derivation decreases, because the gradient of posterior distribution declines and make the
scale of $\Delta'$ deminished when the sampled point is apporching MAP. Thus, the error our
transition kernel will declines with even constant step size as discribed in Sect.~ 4.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{exp2.png}
    \caption{Experiment results of Logistic regression}
    %\vspace{-10pt}
\end{figure}

We use this experiment to illustrate another property of our mini-batch MH testing: when the size of
mini-batch increase to a threshold, the accuracy / log likelihood of sampled data at a fix iteration
number will not improve with larger mini-batch size, see Fig.~4c. This is because the variance of
$\Delta'$ decrease with large mini-batch, and the error of our mini-batch MH testing is proportional
to our estimation error of variance of $\Delta'$. When the $var(\Delta')$ decrease to a threshold,
the error of our mini-batch MH testing can be ignored.

{\color{blue}
Daniel/Haoyu: For additional details, see Appendix~\ref{app:logistic}.
}

@@

In this experiment, we apply about mini-batch MH testing with SGHMC (Stochastic Gradient Hamiltonian Monte Carlo, \cite{sghmc_2014}) proposer to sample the parameter instances from a fully connected neural network. We use Higgs data set from University of California Irvine. It is a classification problem to distinguish between a signal process which produces Higgs bosons and a background process which does not. We build a 4 layers neural network to tackle this problem. The first layer is a input layer, the last layer is a softmax layer, the intermediate layers are linear followed by Sigmoid, norm, dropout, starting and ending with linear operation.

We choose the Adaptive Gradient Descent method ~\cite{adapGrad} as the comparison baseline. Since the frequencies of the features are quite different, we use the same strategy in the Adaptive Gradient to rescale the gradient before apply SGHMC. We implement our system within the open-source BIDMach project~\cite{canny2013bidmach}.

Figure~5 illustrates our experiment results.  
\begin{figure}[t]
    \centering
    \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
    %\includegraphics[width=0.5\linewidth]{empty}
    \caption{This will hopefully show results for neural networks on BIDMach.}
    %\vspace{-10pt}
\end{figure}

@@

{\color{blue}
Daniel: TODO, clean this up.
}

@@

From Lemma~\ref{lem:gaussian}, only the $\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid
\theta_t))$ term brings randomness into $\Delta'$. To simplify the subsequent notation, we define
$g_i(\theta) = \log p(x_i\mid \theta)$, and $q_i = g_i(\theta') - g_i(\theta_t)$. Then by Taylor's Theorem:
\[
|q| = |g(\theta') - g(\theta_t)| = |\nabla g(\theta_t)^T(\theta' - \theta_t) + o(\theta' - \theta_t)^2| \leq 2\epsilon k.
\]
Thus, with given the $\theta$, and $\theta'$, the variance of $\Delta'$ can be write as:
\begin{align*}
{\rm Var}(\Delta') & = {\rm Var}\left[\sum_{i=1}^m g_i(\theta') - g_i(\theta_t)\right] = {\rm Var}(\sum_i^m q_i) = m {\rm Var} (q_i) + m(m-1) {\rm Cov}_{i \neq j}(q_i,q_j) \\
&= m{\rm Var}(q_i) - \frac{m(m-1)}{N-1} {\rm Var}(q_i) \leq 4(m - \frac{m(m-1)}{N-1})k^2 \epsilon^2, \\
\end{align*}
as desired.

>>>>

We conduct three sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression. In
Section~\ref{ssec:nets}, we apply the test for deep learning.
Appendices~\ref{app:gaussian},~\ref{app:logistic}, and~\ref{app:nnet} contain more detailed
information on these experiments.

@@

From the figures above, we find that with the help of our mini-batch MH testing we can obtain the
sampled distribution which are very close to target one. We feature draw our estimation of standard
derivation of $X_{\rm norm}$ over iterations, see Fig.~3e. We find with even constant step size, the
standard derivation decreases, because the gradient of posterior distribution declines and make the
scale of $\Delta'$ deminished when the sampled point is apporching MAP. Thus, the error our
transition kernel will declines with even constant step size as discribed in Section~\ref{sec:theory}.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{exp2.png}
    \caption{Experiment results of Logistic regression}
    \label{fig:logistic_fig}
    %\vspace{-10pt}
\end{figure}

We use this experiment to illustrate another property of our mini-batch MH testing: when the size of
mini-batch increase to a threshold, the accuracy / log likelihood of sampled data at a fix iteration
number will not improve with larger mini-batch size, see Fig.~4c. This is because the variance of
$\Delta'$ decrease with large mini-batch, and the error of our mini-batch MH testing is proportional
to our estimation error of variance of $\Delta'$. When the ${\rm Var}(\Delta')$ decrease to a
threshold, the error of our mini-batch MH testing can be ignored.

Figure~\ref{fig:logistic_fig} shows our results ... {\color{blue} Daniel: TODO finish description}

@@

In this experiment, we apply about mini-batch MH testing with SGHMC (Stochastic Gradient Hamiltonian
Monte Carlo~\cite{sghmc_2014}) proposer to sample the parameter instances from a fully connected
neural network. We use Higgs data set from University of California Irvine. It is a classification
problem to distinguish between a signal process which produces Higgs bosons and a background process
which does not. We build a 4 layers neural network to tackle this problem. The first layer is a
input layer, the last layer is a softmax layer, the intermediate layers are linear followed by
Sigmoid, norm, dropout, starting and ending with linear operation.

We choose the Adaptive Gradient Descent method~\cite{adapGrad} as the comparison baseline. Since
the frequencies of the features are quite different, we use the same strategy in the Adaptive
Gradient to rescale the gradient before apply SGHMC. We implement our system within the open-source
BIDMach project~\cite{canny2013bidmach}.

Figure~\ref{fig:nnet_fig} illustrates our experiment results.

\begin{figure}[t]
    \centering
    \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
    %\includegraphics[width=0.5\linewidth]{empty}
    \caption{This will hopefully show results for neural networks on BIDMach.}
    \label{fig:nnet_fig}
    %\vspace{-10pt}
\end{figure}

@@

{\color{blue}
Daniel: TODO, clean this up!
}

@@

From Lemma~\ref{lem:gaussian}, only the $\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid
\theta_t))$ term brings randomness into $\Delta'$. To simplify the subsequent notation, we define
$g_i(\theta) = \log p(x_i\mid \theta)$, and $q_i = g_i(\theta') - g_i(\theta_t)$. Then by Taylor's Theorem:
\[
|q| = |g(\theta') - g(\theta_t)| = |\nabla g(\theta_t)^T(\theta' - \theta_t) + o(\theta' - \theta_t)^2| \leq 2\epsilon k.
\]
Thus, given $\theta$ and $\theta'$, the variance of $\Delta'$ can be written as:
\begin{align*}
{\rm Var}(\Delta') &= {\rm Var}\left[\sum_{i=1}^m g_i(\theta') - g_i(\theta_t)\right] = {\rm Var}\left(\sum_i^m q_i\right) \\
&= m {\rm Var} (q_i) + m(m-1) {\rm Cov}_{i \neq j}(q_i,q_j) \\
&= m{\rm Var}(q_i) - \frac{m(m-1)}{N-1} {\rm Var}(q_i) \\
&\leq 4\left(m - \frac{m(m-1)}{N-1}\right)k^2 \epsilon^2,
\end{align*}
as desired.
